# Time series decomposition

[The source of the chapter](https://otexts.com/fpp3/decomposition.html)

```{r ch03-ts-decomposition-1}

# loading libraries
library(tsibble)
library(tsibbledata)
library(tidyverse)
# to read data
library(rio)
library(ggplot2)
library(fabletools)
library(feasts)
library(fpp3)
library(latex2exp)
library(forecast)

```

Components:

-   Trend
-   Cyclic
-   Seasonal (some times there can be more than one seasonal component)
-   Remainder component

## Transformations and adjustments

[The source of the section](https://otexts.com/fpp3/transformations.html)

To remove the source of variation or make the pattern core consistent across the data set, the following adjustments and transformation will be considered:

-   calendar adjustments
-   population adjustments
-   inflation adjustments
-   mathematical transformations

### Calendar adjustments

Calendar adjustments of time series are important when dealing with time series data that are influenced by variations in the number of days in a month, public holidays, weekends, or other calendar-related factors. These adjustments help remove calendar effects and make the data more comparable over time.

> For example... It is easy to remove this variation by computing average sales per trading day in each month, rather than total sales in the month.

### Population adjustments

> Any data that are affected by population changes can be adjusted to give per-capita data. That is, consider the data per person (or per thousand people, or per million people) rather than the total.

```{r ch03-ts-decomposition-2}

head(global_economy)

global_economy |>
  filter(Country == "Australia") |>
  autoplot(GDP/Population) +
  labs(title= "GDP per capita", y = "$US")

```

### Inflation adjustments

> To make these adjustments, a price index is used. If $z_t$ denotes the price index and $y_t$ denotes the original house price in year $t$, then $x_t = y_t/z_t \times z_{2000}$ gives the adjusted house price at year 2000 dollar values. Price indexes are often constructed by government agencies. For consumer goods, a common price index is the Consumer Price Index (or CPI).

This allows us to compare the growth or decline of industries relative to a common price value. For example, looking at aggregate annual "newspaper and book" retail turnover from `aus_retail`, and adjusting the data for inflation using CPI from `global_economy` allows us to understand the changes over time.

```{r ch03-ts-decomposition-3, warning=FALSE}

# sum turnover for printing by year
print_retail <- aus_retail |>  # monthly retail turnover in Australia in $M AUD
  filter(Industry == 'Newspaper and book retailing') |> # filter printing
  index_by(Year = year(Month)) |> # grouping by index (key: Month)
  summarise(Turnover = sum(Turnover)) # summarise

# get all economic data by year for Aus
aus_economy <- global_economy |>
  filter(Code == 'AUS')

print_retail |>
  left_join(aus_economy, by='Year') |>
  # Inflation adjustments
  mutate(Adjusted_turnover_base2010 = Turnover / CPI * 100) |>
                                                    # ^^^ base value (2010)
  pivot_longer(cols = c(Turnover, Adjusted_turnover_base2010),
               values_to = 'Turnover') |>
  mutate(name = factor(name, levels = c('Turnover', 'Adjusted_turnover_base2010'))) |>
  ggplot(aes(x=Year, y = Turnover)) +
  geom_line() +
  facet_grid(name ~ ., scales = 'free_y') +
  labs(title = 'Turnover: Australian print media industry',
       y = '$AU')
  


```

> By adjusting for inflation using the CPI, we can see that Australia's newspaper and book retailing industry has been in decline much longer than the original data suggests. The adjusted turnover is in 2010 Australian dollars, as CPI is 100 in 2010 in this data set.

### Mathematical transformations

> If the data shows variation that increases or decreases with the level of the series, then a transformation can be useful.

If we the original observations as $y_1, ..., y_T$ and the transformed observations as $w_1, ..., w_T$, then $w_t = log(y_t)$. Changes in a log value are relative (or percentage) changes on the original scale.

> If any value of the original series is zero or negative, then logarithms are not possible.

#### Box-Cox transformations

$$ w_t = 
\begin{cases}
ln(y_t) & \lambda = 0; \\
\frac {(y_t + 1)^{\lambda}-1} {\lambda} & otherwise
\end{cases} $$

```{r ch03-ts-decomposition-4}

aus_production |>
  autoplot(Gas)

# the optimal lambda
lambda <- BoxCox.lambda(aus_production$Gas)
print(paste('BoxCox lambda:', lambda))

aus_production |>
  mutate(Gas_boxcox = BoxCox(Gas, BoxCox.lambda(Gas))) |>
  autoplot(Gas_boxcox) +
  labs(title = 'BoxCox transformed gas')


aus_production |>
  mutate(Gas_boxcox = box_cox(Gas, lambda)) |>
  mutate(Gas_inv = inv_box_cox(Gas_boxcox, lambda)) |>
  select(Gas, Gas_boxcox, Gas_inv)

```

Choosing $\lambda$.

```{r ch03-ts-decomposition-5}

# way 1 - forecast pkg
print(paste('BoxCox lambda (BoxCox.lambda(...)):', BoxCox.lambda(aus_production$Gas)))
# way 2 - feasts pkg
print(paste('BoxCox lambda (guerrero(...)):', guerrero(aus_production$Gas)))
# way 3 - fabletools pkg
lambda <- aus_production |>
  features(Gas, features = guerrero) |> 
  pull(lambda_guerrero)

print(paste('BoxCox lambda (features(..., features = guerrero)):', lambda))

```

```{r ch03-ts-decomposition-6}

aus_production |>
  autoplot(box_cox(Gas, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed gas production with $\\lambda$ = ",
         round(lambda,2))))

```

## Time series components

[The source of the section](https://otexts.com/fpp3/components.html)

-   **Trend** pattern exists when there is a long-term increase or decrease in the data
-   **Cyclic** pattern exists when data exhibit rises and falls that are not of fixed period (duration usually of at least 2 years)
-   **Seasonal** pattern exists when a series is influenced by seasonal factors (quarter, month, day of the week)

The **additive** decomposition of a time series is represented as:

$$Y_t = T_t + S_t + R_t$$

where:

-   $Y_t$ is the observed value at time $t$,\
-   $T_t$ is the trend component,\
-   $S_t$ is the seasonal component,\
-   $R_t$ is the residual (or random) component.

**Trend** is a smoothly changing component of a time series.

**Seasonality** is a strong periodic component with stable intensity.

The **multiplicative** decomposition of a time series is represented as:

$$Y_t = T_t \times S_t \times R_t$$

where:\
- $Y_t$ is the observed value at time $t$,\
- $T_t$ is the trend component,\
- $S_t$ is the seasonal component,\
- $R_t$ is the residual (or random) component.

Transform additive to multiplicative:

$$ln(Y_t) = ln(T_t \times S_t \times R_t) = ln(T_t) + ln(S_t) + ln(R_t)$$ \### Example: Employment in the US retail sector

```{r ch03-ts-decomposition-7}

us_retail_employment <- us_employment |>
        filter(year(Month) >= 1990, Title == 'Retail Trade') |>
  select(-Series_ID) 

us_retail_employment |>
  autoplot(Employed) +
  labs(title = 'Total employment in US retail',
       y = 'Persons, K')

```

STL decomposition.

```{r ch03-ts-decomposition-8}

dcmp <- us_retail_employment |>
  model(stl = STL(Employed))

dcmp |> components()

```

```{r ch03-ts-decomposition-9}

dcmp |> 
  components() |>
  as_tsibble() |>
  autoplot(Employed, color = 'gray') +
  geom_line(aes(y=trend), color = 'orange') +
  labs(y = 'Persons, K',
       title = 'Total employment in US retail')

```

```{r ch03-ts-decomposition-10}

dcmp |> 
  components() |>
  autoplot()

```

> Notice that the seasonal component changes over time, so that any two consecutive years have similar patterns, but years far apart may have different seasonal patterns.

### Seasonally adjusted data

> If the seasonal component is removed from the original data, the resulting values are the "seasonally adjusted" data. For an additive decomposition, the seasonally adjusted data are given by $y_t − S_t$, and for multiplicative data, the seasonally adjusted values are obtained using $y_t/S_t$.

```{r ch03-ts-decomposition-11}

dcmp |>
  components() |>
  as_tsibble() |> # <<<<<<<<!!!!!!!!!!!
  autoplot(Employed, color = 'gray') +
  geom_line(aes(y=season_adjust), color = 'blue') +
  labs(y = 'Persons, K',
       title = 'Total employment in US retail')
  
```

> If the purpose is to look for turning points in a series, and interpret any changes in direction, then it is better to use the trend-cycle component rather than the seasonally adjusted data.

## Moving averages

[The source of the section](https://otexts.com/fpp3/moving-averages.html)

### Moving average smoothing

$m$-MA, or a moving average of order m:

$$
\hat{T_t} = \frac {1} {m} \sum_{j=-k}^k y_{t+j}, \text{where} \space m = 2k + 1
$$

Moving average removes seasonal fluctuations and allows to estimate trend-cycle.

```{r ch03-ts-decomposition-12}

global_economy |>
  filter(Code == 'AUS') |>
  autoplot(Exports) +
  labs(title = 'Total Australian exports', y = '% of GDP')

```

Computing 5-MA.

```{r ch03-ts-decomposition-13}

# m = 5 => (5 - 1)/2 = 2 = k

global_economy |>
  filter(Code == 'AUS') |>
  mutate(`5-MA` = slider::slide_dbl(Exports, mean, .before = 2, .after = 2, 
                                    .complete = T)) |>
  select(Exports, `5-MA`)

```

```{r ch03-ts-decomposition-14}

global_economy |>
  filter(Code == 'AUS') |>
  mutate(`5-MA` = slider::slide_dbl(Exports, mean, .before = 2, .after = 2, 
                                    .complete = T)) |>
  autoplot(Exports) +
  geom_line(aes(y = `5-MA`), color = 'red') +
  labs(title = 'Total Australian exports', y = '% of GDP')

```

The order of the moving average determines the smoothness of the trend-cycle extimate.

```{r ch03-ts-decomposition-15, warning=FALSE}

global_economy |>
  filter(Code == 'AUS') |>
  mutate(`3-MA` = slider::slide_dbl(Exports, mean, .before = 1, .after = 1, 
                                    .complete = T),
         `5-MA` = slider::slide_dbl(Exports, mean, .before = 2, .after = 2, 
                                    .complete = T),
         `7-MA` = slider::slide_dbl(Exports, mean, .before = 3, .after = 3, 
                                    .complete = T),
         `9-MA` = slider::slide_dbl(Exports, mean, .before = 4, .after = 4, 
                                    .complete = T),
         `11-MA` = slider::slide_dbl(Exports, mean, .before = 5, .after = 5, 
                                    .complete = T),
         `13-MA` = slider::slide_dbl(Exports, mean, .before = 6, .after = 6, 
                                    .complete = T)) |>
  pivot_longer(cols=c(`3-MA`, `5-MA`, `7-MA`, `9-MA`, `11-MA`, `13-MA`),
               values_to = 'ma_value') |>
  mutate(name = factor(name, levels=c('3-MA', '5-MA', '7-MA', 
                                      '9-MA', '11-MA', '13-MA'))) |>
  autoplot(Exports, color = 'gray') +
  geom_line(aes(y = ma_value), color = 'red') +
  labs(title = 'Total Australian exports', y = '% of GDP') +
  facet_wrap(name ~ ., scales = 'free_y', ncol = 2) +
  theme(legend.position = 'none')

```

> This is so they are symmetric: in a moving average of order $m = 2k + 1$, the middle observation, and $k$ observations on either side, are averaged. But if $m$ was even, it would no longer be symmetric.

### Moving averages of moving averages

To make an even-order moving average symmetric, apply a moving average to moving average.

```{r ch03-ts-decomposition-16}

aus_production |>
  filter(year(Quarter) >= 1992) |>
  select(Quarter, Beer) |>
  mutate(`4-MA` = slider::slide_dbl(Beer, mean, .before = 1, .after = 2,  
                                     .complete = TRUE),
         `2x4-MA` = slider::slide_dbl(`4-MA`, mean, .before = 1, .after = 0,  
                                     .complete = TRUE))

```

```{r ch03-ts-decomposition-17}

ma((aus_production |>
  filter(year(Quarter) >= 1992) |>
  select(Quarter, Beer))$Beer, 4)

```

$2 \times m$-MA:

$$
\hat{T_t} = \frac {1} {2} \left[ \frac {1} {m} \sum_{j=-k}^k y_{t+j} \right], \text{where} \space m = 2k
$$

> In general, an even order MA should be followed by an even order MA to make it symmetric. Similarly, an odd order MA should be followed by an odd order MA.

```{r ch03-ts-decomposition-18}

aus_production |>
  filter(year(Quarter) >= 1992) |>
  select(Quarter, Beer) |>
  mutate(`3-MA` = slider::slide_dbl(Beer, mean, .before = 1, .after = 1,  
                                     .complete = TRUE),
         `3x3-MA` = slider::slide_dbl(`3-MA`, mean, .before = 1, .after = 1,  
                                     .complete = TRUE))

```

### Estimating the trend-cycle with seasonal data

> If the seasonal period is even and of order $m$, we use a $2 \times m$-MA to estimate the trend-cycle. If the seasonal period is odd and of order $m$, we use a $m$-MA to estimate the trend-cycle. For example, a 2×12-MA can be used to estimate the trend-cycle of monthly data with annual seasonality and a 7-MA can be used to estimate the trend-cycle of daily data with a weekly seasonality.

### Example: Employment in the US retail sector

```{r ch03-ts-decomposition-19, warning=FALSE}

us_retail_employment |>
  gg_season(Employed)

us_retail_employment |>
  mutate(`12-MA` = slider::slide_dbl(Employed, mean, .before=5,
                                     .after = 6, .complete = TRUE),
         `2x12-MA` = slider::slide_dbl(`12-MA`, mean, .before=1,
                                     .after = 0, .complete = TRUE)) |>
autoplot(Employed, color = 'gray') +
  geom_line(aes(y=`2x12-MA`), color = 'red') +
   labs(y = "Persons (thousands)",
       title = "Total employment in US retail")

```

### Weighted moving averages

$$
\hat{T_t} = \sum_{j=-k}^k a_j y_{t+j}, \text{where} \space k = (m - 1)/2
$$

and the weights are given by $[a_{−k}, ...,a_k]$. It is important that the weights all sum to one and that they are symmetric so that $a_j = a_{−j}$. The simple $m$-MA is a special case where all of the weights are equal to $1/m$.

> A major advantage of weighted moving averages is that they yield a smoother estimate of the trend-cycle.

## Classical decomposition

[The source of the section](https://otexts.com/fpp3/classical-decomposition.html)

Classical decomposition:

-   Additive decomposition
-   Multiplicative decomposition

### Additive decomposition

-   **Step 1**\
    Compute $\hat T_t$ using:

    -   $2 \times m$-MA, if $m$ is an even number
    -   $m$-MA, if $m$ is an odd number

-   **Step 2**\
    Calculate the detrended series: $y_t - \hat T_t$

-   **Step 3**\
    To estimate the seasonal component $\hat S_t$ for each season, simply average the detrended values for that season.

-   **Step 4**\
    $$
    \hat R_t = y_t - \hat T_t - \hat S_t
    $$

```{r ch03-ts-decomposition-20}
# classical decomposition

us_retail_employment |>
  model(classical_decomposition(Employed, type='add')) |>
  components() |>
  autoplot() +
  labs(title = "Classical additive decomposition of total
                  US retail employment")

```

### Multiplicative decomposition

-   **Step 1**\
    Compute $\hat T_t$ using:

    -   $2 \times m$-MA, if $m$ is an even number
    -   $m$-MA, if $m$ is an odd number

-   **Step 2**\
    Calculate the detrended series: $y_t / \hat T_t$

-   **Step 3**\
    To estimate the seasonal component $\hat S_t$ for each season, simply average the detrended values for that season.

-   **Step 4**\
    $$
    \hat R_t = y_t /( \hat T_t \hat S_t)
    $$ \## STL decomposition

[The source of the section](https://otexts.com/fpp3/stl.html)

> STL decomposition only provides facilities for additive decompositions. A multiplicative decomposition can be obtained by first taking logs of the data, then back-transforming the components. Decompositions that are between additive and multiplicative can be obtained using a Box-Cox transformation of the data with $0 < \lambda < 1$. A value of \$\lambda = 0 \$ gives a **multiplicative decomposition** while $\lambda = 1$ gives an **additive decomposition**.

```{r ch03-ts-decomposition-21}

us_retail_employment |>
  model(stl = STL(Employed ~ trend(window = 7) +
                    season(window = 'periodic'), robust = TRUE)) |>
  components()

```

```{r ch03-ts-decomposition-22}

us_retail_employment |>
  model(stl = STL(Employed ~ trend(window = 7) +
                    season(window = 'periodic'), robust = TRUE)) |>
  components() |>
  autoplot()

```

The main parameters:

-   `trend(window = ?)` controls how rapidly the trend-cycle components can change.
-   `season(window = ?)` controls how rapidly the seasonal components can change.

> Both trend and seasonal windows should be **odd** numbers. Setting the seasonal window to be infinite is equivalent to forcing the seasonal component to be periodic.

## Excercises

1.  

```{r ch03-ts-decomposition-23}
# Plot the GDP per capita for each country over time.
for(country in (global_economy |> distinct(Country))$Country){
  gdp <- global_economy |>
  filter(Country == country) |>
  mutate(GDP_per_capita = GDP / Population) |>
  autoplot(GDP_per_capita) +
  facet_grid(Country ~ ., scales = 'free_y') +
  labs(title = 'GDP per capita', subtitle = country, y =  ' $USD per capita') +
  theme(legend.position = 'none')
  print(gdp)
}

```

```{r ch03-ts-decomposition-24}

#Which country has the highest GDP per capita?

highest_countries <- (global_economy |>
                  mutate(GDP_per_capita = GDP / Population) |>
                  filter(Year == max(global_economy$Year)) |>
                  arrange(desc(GDP_per_capita)) |>
                  select(Country, GDP_per_capita) |>
                  head(10))$Country

for(country in highest_countries){
  gdp <- global_economy |>
  filter(Country == country) |>
  mutate(GDP_per_capita = GDP / Population) |>
  autoplot(GDP_per_capita) +
  facet_grid(Country ~ ., scales = 'free_y') +
  labs(title = 'GDP per capita', subtitle = paste('#', which(highest_countries == country, arr.ind = TRUE), country), y =  ' $USD per capita') +
  theme(legend.position = 'none')
  print(gdp)
}

```

2.  

```{r ch03-ts-decomposition-25}

# United States GDP from global_economy.
global_economy |>
  filter(Country == 'United States') |>
  autoplot(GDP) +
  labs(title = 'GDP of United States', y = '$US')

# => transformation does not needed

```

```{r ch03-ts-decomposition-26}

# Slaughter of Victorian “Bulls, bullocks and steers” in aus_livestock.

aus_livestock |>
  filter(State == 'Victoria', Animal == 'Bulls, bullocks and steers') |>
  autoplot(Count) +
  labs(title = 'Slaughter of Bulls, bullocks and steers', subtitle = 'Victoria')

# => Apply transformation to make variance constant

bulls <- aus_livestock |>
  filter(State == 'Victoria', Animal == 'Bulls, bullocks and steers')

print(paste('Optimal lambda: ', BoxCox.lambda(bulls$Count)))

bulls |>
  mutate(Count_boxcox = box_cox(Count, lambda=BoxCox.lambda(bulls$Count))) |>
  autoplot(Count_boxcox) +
  labs(title = 'Slaughter of Bulls, bullocks and steers (BoxCox)', subtitle = 'Victoria')

# => Variance became more constant

```

```{r ch03-ts-decomposition-27}

# Victorian Electricity Demand from vic_elec.

vic_elec |>
  autoplot(Demand) +
  labs(title = 'Victorian Electricity Demand', subtitle = 'Victoria', y='GWh')

# apply transformation

lambda <- vic_elec |>
  features(Demand, features = guerrero) |> 
  pull(lambda_guerrero)

vic_elec |>
  mutate(Demand_cb = BoxCox(Demand, lambda = lambda)) |>
  autoplot(Demand_cb) +
  labs(title = 'Victorian Electricity Demand (BoxCox)', subtitle = 'Victoria', y='GWh')

# => Peaks became lower

```

```{r ch03-ts-decomposition-28}

# Gas production from aus_production.

aus_production |>
  autoplot(Gas) +
  labs(y = "",
       title = "Gas production in Australia")

# => Apply transformation

lambda <- aus_production |>
  features(Gas, features = guerrero) |> 
  pull(lambda_guerrero)

print(paste('Optimal BoxCox lambda:', lambda))

aus_production |>
  autoplot(Gas, color = 'gray') +
  geom_line(aes(y=box_cox(Gas, lambda)), color='red') +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed gas production with $\\lambda$ = ",
         round(lambda,2))))

# => Variance became more constant

```

3.  

```{r ch03-ts-decomposition-29}

canadian_gas |>
  autoplot(Volume) +
  labs(title = 'Monthly Canadian gas production', y = 'Billions of cubic metres')

lambda <- canadian_gas |>
  features(Volume, features = guerrero) |> 
  pull(lambda_guerrero)

canadian_gas |>
  mutate(Volume_bc = box_cox(Volume, lambda=lambda)) |>
  autoplot(Volume, color='gray') +
  geom_line(aes(y=Volume_bc), color='blue') +
  labs(title = 'Monthly Canadian gas production', y = 'Billions of cubic metres')

# => BoxCox is unhelpful due to seasonality

canadian_gas |>
  model(stl = STL(Volume ~ trend(window=21) + season(window='periodic'))) |>
  components() |>
  autoplot()

```

4.  

```{r ch03-ts-decomposition-30}

myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))

lambda <- myseries |>
  features(Turnover, features = guerrero) |> 
  pull(lambda_guerrero)

myseries |>
  autoplot(box_cox(Turnover, lambda))

```

5.  

```{r ch03-ts-decomposition-31, warning=FALSE}

# Tobacco from aus_production

lambda <- aus_production |>
  features(Tobacco, features = guerrero) |> 
  pull(lambda_guerrero)

aus_production |>
  autoplot(Tobacco, color = 'gray') +
  geom_line(aes(y = box_cox(Tobacco, lambda=lambda)), color='blue') +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed tobacco production with $\\lambda$ = ",
         round(lambda,2))))

```

```{r ch03-ts-decomposition-32, warning=FALSE}

# Economy class passengers between Melbourne and Sydney from ansett

pass_mel_syd <- ansett |>
  filter(Class == 'Economy', Airports == 'MEL-SYD')

lambda <- pass_mel_syd |>
  features(Passengers, features = guerrero) |> 
  pull(lambda_guerrero)

pass_mel_syd  |>
  mutate(Pass_bc = box_cox(Passengers, lambda)) |>
  autoplot(Pass_bc, color='blue') +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed a number of passangers with $\\lambda$ = ",
         round(lambda,2))),
       subtitle = 'Economy class passengers between Melbourne and Sydney')
  
```

```{r ch03-ts-decomposition-33}

# Pedestrian counts at Southern Cross Station from pedestrian

south_cross <- pedestrian |>
  filter(Sensor == 'Southern Cross Station')

lambda <- south_cross |>
  features(Count, features = guerrero) |> 
  pull(lambda_guerrero)

south_cross  |>
  mutate(Count_bc = box_cox(Count, lambda)) |>
  autoplot(Count_bc, color='blue') +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed a number of pedestrians with $\\lambda$ = ",
         round(lambda,2))),
       subtitle = 'Southern Cross Station')

```

6.  

```{r ch03-ts-decomposition-34}

df <- data.frame(value = sample.int(7))
w <- c(0.067, 0.133, 0.200, 0.200, 0.200, 0.133, 0.067)

print(paste('3x5-MA = ', 
            round((df |>
                  mutate(`5-MA` = slider::slide_dbl(value, mean, 
                                                    .before = 2, 
                                                    .after = 2,
                                                    .complete = TRUE),
                         `3x5-MA` = slider::slide_dbl(`5-MA`, mean, 
                                                      .before = 1, 
                                                      .after = 1,
                                                    .complete = TRUE)) |>
                  filter(!is.na(`3x5-MA`)))$`3x5-MA`, 2),
      ', 7-term weighted MA = ', 
      round((cbind(df, w) |>
        summarise(`7-term MA` = sum(value * w)))$`7-term MA`, 2)))

```

7.  

```{r ch03-ts-decomposition-35}

gas <- tail(aus_production, 5*4) |> select(Gas)

# a.

gas |>
  autoplot()

# => Seasonal fluctuations: summer - less gas production, winter - more, 
#                           trend-cycle - increasing trend

```

```{r ch03-ts-decomposition-36}

# b.

c_dec <- gas |>
          model(c_dec = classical_decomposition(Gas, type='mul')) |>
          components()

c_dec |>
  autoplot()

# c. 

# Do the results support the graphical interpretation from part a? - YES

```

```{r ch03-ts-decomposition-37}

# d.

c_dec |>
  autoplot(season_adjust)

```

```{r ch03-ts-decomposition-38}

# e.

gas$Gas[length(gas$Gas) / 2] = 300

gas |>
      model(c_dec = classical_decomposition(Gas, type='mul')) |>
      components() |>
      autoplot(season_adjust)

```

```{r ch03-ts-decomposition-39}

# f.

gas <- tail(aus_production, 5*4) |> select(Gas)
gas$Gas[length(gas$Gas)] = 300

gas |>
      model(c_dec = classical_decomposition(Gas, type='mul')) |>
      components() |>
      autoplot(season_adjust)

# => the outlier in the middle of the time series changes seasonal pattern

```

8.  

```{r ch03-ts-decomposition-40}

# X-11 does not work properly on the MAC OS (in my case)

```

9.  

    a.  The decomposition of the Australian quarterly labor force data into trend, seasonal, and remainder components reveals distinct patterns. In the first plot (Figure 3.19), the trend component shows a steady increase in the labor force over time, with a slight dip around the 1991/1992 period, indicating the effects of the recession. The seasonal component reflects regular fluctuations across the quarters, but these fluctuations are relatively small in magnitude, suggesting that the seasonal variation in labor force participation is limited compared to the overall trend. In the second plot (Figure 3.20), which displays the seasonal component, we can observe that the recession of 1991/1992 is indeed visible as a notable deviation from the baseline, particularly in the downward spikes during that period.
    b.  The scale of the reminder component is much smaller than the trend, but the recession's impact is clearly captured as a short-term anomaly, confirming the influence of the economic downturn on the labor market. Overall, the recession is most visible in the trend and remainder components.

10. 

```{r ch03-ts-decomposition-41}

# a.
canadian_gas |>
  autoplot()

canadian_gas |>
  gg_season()

canadian_gas |>
  gg_subseries()


```

```{r ch03-ts-decomposition-42}

# b.
canadian_gas |>
  model(stl = STL(Volume ~ trend(window=11) + season(window=21))) |>
  components() |>
  autoplot()

```

```{r ch03-ts-decomposition-43}

# c.

canadian_gas |>
  gg_season()

```

```{r ch03-ts-decomposition-44}

#d.

canadian_gas |>
  model(stl = STL(Volume ~ trend(window=11) + season(window=21))) |>
  components() |>
  autoplot(season_adjust)

```
