# Time series features

[The source of the chapter](https://otexts.com/fpp3/features.html)

```{r ch04-ts-features-1}

# loading libraries
library(tsibble)
library(tsibbledata)
library(tidyverse)
# to read data
library(rio)
library(ggplot2)
library(fabletools)
library(feasts)
library(fpp3)
library(latex2exp)
library(forecast)

```

Different features of time series could be used to explore the properties of the series.

## Some simple statistics

[The source of the section](https://otexts.com/fpp3/some-simple-statistics.html)

```{r ch04-ts-features-2}

tourism |>
  features(Trips, list(mean=mean)) |>
  arrange(mean)

```

```{r ch04-ts-features-3}

tourism |>
  features(Trips, quantile)

```

### ACF features

[The source of the section](https://otexts.com/fpp3/acf-features.html)

```{r ch04-ts-features-4}

tourism |>
  features(Trips, feat_acf)

```

### STL features

[The source of the section](https://otexts.com/fpp3/stlfeatures.html)

1.  Strength of trend:

> For strongly trended data, the seasonally adjusted data should have much more variation than the remainder But for data with little or no trend, the two variances should be approximately the same.

$$
F_T = max(0, 1 - \frac {Var(R_t)} {Var(T_t + R_t)}), [0, 1]
$$

2.  Strength of seasonality:

> A series with seasonal strength $F_S$ close to 0 exhibits almost no seasonality, while a series with strong seasonality will have $F_S$ close to 1.

$$
F_S = max(0, 1 - \frac {Var(R_t)} {Var(S_t + R_t)}), [0, 1]
$$

```{r ch04-ts-features-5}

tourism |>
  features(Trips, feat_stl)

```

To identify what type of series are heavily trended and what are most seasonal.

```{r ch04-ts-features-6}

tourism |>
  features(Trips, feat_stl) |>
  ggplot(aes(x = trend_strength, y = seasonal_strength_year, col = Purpose)) +
  geom_point() +
  facet_wrap(vars(State))


```

The most seasonal series can also be easily identified and plotted.

```{r ch04-ts-features-7}

tourism |>
  features(Trips, feat_stl) |>
  filter(seasonal_strength_year == max(seasonal_strength_year)) |>
  left_join(tourism, by=c('State', 'Region', 'Purpose'), multiple='all') |>
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  facet_wrap(vars(State, Region, Purpose))

```

`feat_stl()` returns:

-   `trend_strength` is strength of trend
-   `seasonal_strength_year` is strength of seasonality
-   `seasonal_peak_year` indicates the timing of the peaks
-   `seasonal_trough_year` indicates the timing of the troughs
-   `spikiness` measures the prevalence of spikes in the remainder component
-   `linearity` measures the linearity of the trend component
-   `curvature` measures the curvature of the trend component
-   `stl_e_acf1` is the first autocorrelation coefficient
-   `stl_e_acf10` is the sum of squares of the first ten autocorrelation coefficients


### Other features

[The source of the section](https://otexts.com/fpp3/other-features.html)

> `coef_hurst` will calculate the Hurst coefficient of a time series which is a measure of “long memory”. A series with long memory will have significant autocorrelations for many lags.

```{r ch04-ts-features-8}

tourism |>
  features(Trips, coef_hurst)

```

__Interpretation:__

+ $H = 0.5$: The time series is a random walk with no memory (random behavior).
+ $H < 0.5$: The time series tends to reverse direction.
+ $H > 0.5$: The time series is persistent, meaning that if the series has been increasing, it is more likely to continue increasing, and if it has been decreasing, it is more likely to continue decreasing.

> `feat_spectral` will compute the (Shannon) spectral entropy of a time series, which is a measure of how easy the series is to forecast. A series which has strong trend and seasonality (and so is easy to forecast) will have entropy close to 0. A series that is very noisy (and so is difficult to forecast) will have entropy close to 1.

```{r ch04-ts-features-9}

tourism |>
  features(Trips, feat_spectral)

```

> `box_pierce` gives the Box-Pierce statistic for testing if a time series is white noise, and the corresponding p-value.

```{r ch04-ts-features-10}

tourism |>
  features(Trips, box_pierce)

```

> `ljung_box` gives the Ljung-Box statistic for testing if a time series is white noise, and the corresponding p-value.

```{r ch04-ts-features-11}

tourism |>
  features(Trips, ljung_box)

```

> `guerrero` computes the optimal $\lambda$ value for a Box-Cox transformation using the Guerrero method

```{r ch04-ts-features-12}

tourism |>
  features(Trips, guerrero)

```


## Exploring Australian tourism data

[The source of the section](https://otexts.com/fpp3/exploring-australian-tourism-data.html)

```{r ch04-ts-features-13}

tourism_features <- tourism |>
  features(Trips, feature_set(pkgs = "feasts"))
tourism_features

```

> Provided the urca and fracdiff packages are installed, this gives 48 features for every combination of the three key variables (Region, State and Purpose). We can treat this tibble like any data set and analyse it to find interesting observations or groups of observations.

```{r ch04-ts-features-14, message=FALSE, warning=FALSE, fig.height=10}

library(glue)
tourism_features |>
  select_at(vars(contains("season"), Purpose)) |>
  mutate(
    seasonal_peak_year = seasonal_peak_year +
      4*(seasonal_peak_year==0),
    seasonal_trough_year = seasonal_trough_year +
      4*(seasonal_trough_year==0),
    seasonal_peak_year = glue("Q{seasonal_peak_year}"),
    seasonal_trough_year = glue("Q{seasonal_trough_year}"),
  ) |>
  GGally::ggpairs(mapping = aes(colour = Purpose))

```

Using principal component method to reduce dimention and explore the variables.

```{r ch04-ts-features-15}

library(broom)
pcs <- tourism_features |>
  select(-State, -Region, -Purpose) |>
  prcomp(scale = TRUE) |>
  augment(tourism_features)
pcs |>
  ggplot(aes(x = .fittedPC1, y = .fittedPC2, col = Purpose)) +
  geom_point() +
  theme(aspect.ratio = 1)

```


Explore outliers.

```{r ch04-ts-features-16}

outliers <- pcs |>
  filter(.fittedPC1 > 10) |>
  select(Region, State, Purpose, .fittedPC1, .fittedPC2)
outliers

```

```{r ch04-ts-features-17}

outliers |>
  left_join(tourism, by = c("State", "Region", "Purpose"), multiple = "all") |>
  mutate(Series = glue("{State}", "{Region}", "{Purpose}", .sep = "\n\n")) |>
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  facet_grid(Series ~ ., scales = "free") +
  labs(title = "Outlying time series in PC space")

```

## Exercises

1. 

```{r ch04-ts-features-18}

?PBS

mean_sd <- function(x){
  c(mean=mean(x), sd = sd(x))
}

PBS |>
  features(Cost, mean_sd) |>
  filter(mean == max(mean)) |>
  left_join(PBS, by = c('Concession', 'Type', 'ATC1', 'ATC2'), multiple = "all") |>
  as_tsibble(index = Month, key = c('Concession', 'Type', 'ATC1', 'ATC2')) |>
  autoplot(Cost) +
  labs(title = 'Time series with the highest mean')

```

2.

```{r ch04-ts-features-19, message=FALSE, warning=FALSE, fig.height=10}

tourism |>
  filter(Purpose == 'Holiday') |>
  features(Trips, feature_set(pkgs = "feasts")) |>
  select_at(vars(contains("season"), State)) |>
  mutate(
    seasonal_peak_year = seasonal_peak_year +
      4*(seasonal_peak_year==0),
    seasonal_trough_year = seasonal_trough_year +
      4*(seasonal_trough_year==0),
    seasonal_peak_year = glue("Q{seasonal_peak_year}"),
    seasonal_trough_year = glue("Q{seasonal_trough_year}"),
  ) |>
  mutate(
    seasonal_peak_year = as.factor(seasonal_peak_year),
    seasonal_trough_year = as.factor(seasonal_trough_year)
  ) |>
  GGally::ggpairs(mapping = aes(colour = State),
                  columns = c("seasonal_peak_year", 
                              "seasonal_trough_year", 
                              "seasonal_strength_year", 
                              "State"))

```


