# Advanced forecasting methods

[The source of the chapter](https://otexts.com/fpp3/advanced.html)

```{r ch12-ts-advanced-1}

# loading libraries
library(tsibble)
library(tsibbledata)
library(tidyverse)
# to read data
library(rio)
library(ggplot2)
library(fabletools)
library(feasts)
library(fpp3)
library(latex2exp)
library(forecast)

```

## Complex seasonality

[The source of the section](https://otexts.com/fpp3/complexseasonality.html)

Hourly data usually has three types of seasonality: a daily pattern, a weekly pattern, and an annual pattern.

```{r ch12-ts-advanced-2}

bank_calls |>
  fill_gaps() |>
  autoplot(Calls) +
  labs(y = "Calls",
       title = "Five-minute call volume to bank")

```

There is complex seasonality in the data:

1. A strong daily seasonal pattern (169 5-minut intervals per day)
2. A weak weekly seasonal pattern with period 169 * 5 = 845
3. If a longer series of data were available, we may also have observed an annual seasonal pattern.


### STL with multiple seasonal periods

```{r ch12-ts-advanced-3}

calls <- bank_calls |>
  mutate(t = row_number()) |>
  update_tsibble(index = t, regular = TRUE)

calls |>
  model(
    STL(sqrt(Calls) ~ season(period = 169) +
                      season(period = 5*169),
        robust = TRUE)
  ) |>
  components() |>
  autoplot() + labs(x = "Observation")

```

```{r ch12-ts-advanced-4}

# Forecasts from STL+ETS decomposition
my_dcmp_spec <- decomposition_model(
  STL(sqrt(Calls) ~ season(period = 169) +
                    season(period = 5*169),
      robust = TRUE),
  ETS(season_adjust ~ season("N"))
)
fc <- calls |>
  model(my_dcmp_spec) |>
  forecast(h = 5 * 169)

# Add correct time stamps to fable
fc_with_times <- bank_calls |>
  new_data(n = 7 * 24 * 60 / 5) |>
  mutate(time = format(DateTime, format = "%H:%M:%S")) |>
  filter(
    time %in% format(bank_calls$DateTime, format = "%H:%M:%S"),
    wday(DateTime, week_start = 1) <= 5
  ) |>
  mutate(t = row_number() + max(calls$t)) |>
  left_join(fc, by = "t") |>
  as_fable(response = "Calls", distribution = Calls)

# Plot results with last 3 weeks of data
fc_with_times |>
  fill_gaps() |>
  autoplot(bank_calls |> tail(14 * 169) |> fill_gaps()) +
  labs(y = "Calls",
       title = "Five-minute call volume to bank")

```

## Prophet model

[The source of the section](https://otexts.com/fpp3/prophet.html)

```{r ch12-ts-advanced-5}

library(fable.prophet)

```
The model for daily data with weekly and yearly seasonality + holiday effects.

$$
y_t = g(t) + s(t) + h(t) + \varepsilon (t)
$$

where

+ $g(t)$ describes a piecewise-linear trend. The knots are automatically selected if not specified.
+ $s(t)$ is the seasonal components that consists of Foyrier terms. By default, the order of period 10 is used for annual seasonality, order 3 is used for weekly.
+ $h(t)$ is holiday effects are added as dummy vars.
+ $\varepsilon (t)$ is a white noise error terms.

### Example: Quarterly cement production

```{r ch12-ts-advanced-6}

cement <- aus_production |>
  filter(year(Quarter) >= 1988)

train <- cement |>
  filter(year(Quarter) <= 2007)

fit <- train |>
  model(
    arima = ARIMA(Cement),
    ets = ETS(Cement),
    prophet = prophet(Cement ~ season(period=4, order = 2, 
                                      type = 'multiplicative'))
  )

```

> **Note!** The seasonal term must have the `period` for quarterly and monthly data.

```{r ch12-ts-advanced-7}

fc <- fit |> forecast(h = "2 years 6 months")
fc |> autoplot(cement)

```

```{r ch12-ts-advanced-8}

fc |> accuracy(cement)

```

### Example: Half-hourly electricity demand

```{r ch12-ts-advanced-9}

vic_elec |>
  pivot_longer(Demand:Temperature, names_to = "Series") |>
  ggplot(aes(x = Time, y = value)) +
  geom_line() +
  facet_grid(rows = vars(Series), scales = "free_y") +
  labs(y = "")

```

```{r ch12-ts-advanced-10}

elec <- vic_elec |>
  mutate(
    DOW = wday(Date, label = TRUE),
    Working_Day = !Holiday & !(DOW %in% c("Sat", "Sun")),
    Cooling = pmax(Temperature, 18)
  )
elec |>
  ggplot(aes(x=Temperature, y=Demand, col=Working_Day)) +
  geom_point(alpha = 0.6) +
  labs(x="Temperature (degrees Celsius)", y="Demand (MWh)")

```

```{r ch12-ts-advanced-11}

fit <- elec |>
  model(
    prophet(Demand ~ Temperature + Cooling + Working_Day +
            season(period = "day", order = 10) +
            season(period = "week", order = 5) +
            season(period = "year", order = 3))
  )
fit |>
  components() |>
  autoplot()

```

```{r ch12-ts-advanced-12}

fit |> gg_tsresiduals()

```

```{r ch12-ts-advanced-13}

elec_newdata <- new_data(elec, 2*48) |>
  mutate(
    Temperature = tail(elec$Temperature, 2 * 48),
    Date = lubridate::as_date(Time),
    DOW = wday(Date, label = TRUE),
    Working_Day = (Date != "2015-01-01") &
                   !(DOW %in% c("Sat", "Sun")),
    Cooling = pmax(Temperature, 18)
  )

fc <- fit |>
  forecast(new_data = elec_newdata)

fc |>
  autoplot(elec |> tail(10 * 48)) +
  labs(x = "Date", y = "Demand (MWh)")

```

## Vector autoregressions

[The source of the section](https://otexts.com/fpp3/VAR.html)


